---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(tidyverse)
```

# washopenresearch

<!-- badges: start -->

<!-- badges: end -->

The goal of washopenresearch is to provide an overview of open research data related to Water Sanitation and Hygiene (WASH). The current version contains two datasets from the following sources:

-   `washdev`: Open access journal [*Journal of Water, Sanitation and Hygiene for Development*](https://iwaponline.com/washdev)
-   `uncnewsletter`: Research section of the newsletter [North Carolina Water News](https://waterinstitute.unc.edu/our-work/nc-water-news-newsletter)

TODO: A plot here

## Installation

You can install the development version of washopenresearch from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/washopenresearch")
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

```{r, echo=FALSE}

extdata_path <- "https://github.com/openwashdata/washopenresearch/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data
The package provides access to two datasets `washdev` and `uncnewsletter`. Each dataset collects information on scientific articles about (1) article metadata (e.g. title, first author, correspondence author), (2)supplementary material information, (3) data availability statement, and (4) semantic information (e.g. keywords)

```{r}
library(washopenresearch)
```

### washdev
The dataset `washdev` contains data on open access articles of the *Journal of Water, Sanitation & Hygiene for Development* (Vol.1 Issue 1 - Vol.13 Issue 11). It has 924 observations from March 2011 to November 2023.

### uncnewsletter
The dataset `uncnewsletter` contains data on a curated list of articles published at the Research section of the newsletter North Carolina Water News. 

## Example

```{r example}
library(washopenresearch)
## basic example code
```

## License

Data are available as [CC-BY](https://github.com/openwashdata/wasteskipsblantyre/blob/main/LICENSE.md).

## Citation
Please cite this package using:

```{r}
citation("washopenresearch")
```

